Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-10-04 MS/ms.Rmd
Removed sentence in figure legend. Fixed first paragraph of new section.

Certain types of heterogeneity, however, get in the way of understanding real biological patterns and are simply a 'nuisance' to interpreting effects. In other words, if we compare different varieties of 'apples', is it still fair to synthesize these studies to understand overall effects? In comparative physiology, this situation might be analogous to a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not of primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth.

@@

The key features of all effect sizes used in meta-analyses are that: 1) they are statistical parameters that have been placed on a common scale so that they can be compared across studies and 2) they have some associated measure of precision (i.e., sampling variance) [@Koricheva2013; @Borenstein2009; @Schmid2021]. Effect sizes in comparative physiology are quite diverse; they can be the mean difference between two groups [e.g., difference in mean between a control and treatment group; e.g., @WuSeebacher2020], the slope of a regression [e.g., between body size and metabolic rate, @Uyeda2017], a correlation between two variables [(]e.g., between hormone levels and behaviour, @Holtman2016], or even the raw mean (or variance) itself [e.g., comparing CTmax or CTmin across species - e.g., @Sunday2014]. In many cases, different effect sizes might be used depending on the question. For example, to investigate the effect of temperature on a physiological trait, one can compile studies experimentally manipulating temperature and analyse the mean 'effect' by creating contrast-based effect sizes, or alternatively, just model the trait in each temperature treatment. The choice as to which one to use boils down to the types of data available from primary studies, the specific question of interest, and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio, *$\ln{RR}$*. Instead, if we are interested in how the mean of some physiological trait changes with temperature, we may model the mean itself. In the latter scenario, as long as all the means are measured in the same units (e.g., degrees Celsius) or can be converted to the same units, then the mean may provide the desired answer to the original question.

@@

```{r, figsurv, fig.wid = 16.833333, fig.height=5,fig.cap="**Summary of meta-analytic practices in comparative physiology.** (A) Different types of effect measures commonly used in meta-analyses and their relative frequency of use. (B) Frequency of studies where ‘nuisance heterogenity’ had the potential to impact effect sizes (x-axis), and the frequencies of studies explicitly dealing with nuisance heterogeneity (bar colour). (C) Frequencies of analytical approaches used to analyse effect measures. 'Orange' bars indicate weighted models. The results of this survey were drawn from 62 studies in comparative physiology."}
p1 = dat %>% dplyr::group_by(effect_size_type) %>% janitor::tabyl(effect_size_type)  %>% 
  dplyr::summarise(SMD = dplyr::if_else(grepl("SMD", effect_size_type), n, 0),
                  lnRR = dplyr::if_else(grepl("Response Ratio", effect_size_type), n, 0),
                    Zr = dplyr::if_else(grepl("correlation", effect_size_type), n, 0),
                 Other = dplyr::if_else(grepl("other|Q10", effect_size_type), n, 0),
                 Mean = dplyr::if_else(grepl("mean/proportion/raw", effect_size_type), n, 0),
                 lnCVR = dplyr::if_else(grepl("lnCVR", effect_size_type), n, 0),
                 Slope = dplyr::if_else(grepl("slope", effect_size_type), n, 0)) %>% 
  colSums(.) %>% data.frame(effect = names(.), n = .) %>% mutate(prop = n / sum(n)) %>% arrange(effect) %>% 
  mutate(effect = fct_relevel(effect, c("Other", "Zr", "lnRR", "SMD", "lnCVR", "Slope", "Mean"))) %>% 
  ggplot() + tm + 
  geom_col(aes(x = effect, y = prop), width = 0.5, fill = "black", colour = text.colour, size = 0.5) +
  scale_y_continuous(expand=c(0,0), limits = c(0,0.35)) +
  labs(x = "Effect Size",
       y = "Frequency") +
  coord_flip()

>>>>

Certain types of heterogeneity get in the way of understanding real biological patterns and are simply a ‘nuisance’ to interpreting effects. Take for example a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall diet effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not of primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Such heterogeneity is an important  contributing factor to the ‘apples and oranges’ problems described above.

@@

The key features of all effect sizes used in meta-analyses are that: 1) they are statistical parameters that have been placed on a common scale so that they can be compared across studies and 2) they have some associated measure of precision (i.e., sampling variance) [@Koricheva2013; @Borenstein2009; @Schmid2021]. Effect sizes in comparative physiology are quite diverse; they can be the mean difference between two groups [e.g., difference in mean between a control and treatment group; e.g., @WuSeebacher2020], the slope of a regression [e.g., between body size and metabolic rate, @Uyeda2017], a correlation between two variables [e.g., between hormone levels and behaviour, @Holtman2016], or even the raw mean (or variance) itself [e.g., comparing CTmax or CTmin across species - e.g., @Sunday2014]. In many cases, different effect sizes might be used depending on the question. For example, to investigate the effect of temperature on a physiological trait, one can compile studies experimentally manipulating temperature and analyse the mean 'effect' by creating contrast-based effect sizes, or alternatively, just model the trait in each temperature treatment. The choice as to which one to use boils down to the types of data available from primary studies, the specific question of interest, and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio, *$\ln{RR}$*. Instead, if we are interested in how the mean of some physiological trait changes with temperature, we may model the mean itself. In the latter scenario, as long as all the means are measured in the same units (e.g., degrees Celsius) or can be converted to the same units, then the mean may provide the desired answer to the original question.

@@

```{r, figsurv, fig.wid = 16.833333, fig.height=5,fig.cap="**Summary of meta-analytic practices in comparative physiology.** (A) Different types of effect measures commonly used in meta-analyses and their relative frequency of use. (B) Frequency of studies where ‘nuisance heterogenity’ had the potential to impact effect sizes (x-axis), and the frequencies of studies explicitly dealing with nuisance heterogeneity (bar colour). (C) Frequencies of analytical approaches used to analyse effect measures. 'Orange' bars indicate weighted models."}
p1 = dat %>% dplyr::group_by(effect_size_type) %>% janitor::tabyl(effect_size_type)  %>% 
  dplyr::summarise(SMD = dplyr::if_else(grepl("SMD", effect_size_type), n, 0),
                  lnRR = dplyr::if_else(grepl("Response Ratio", effect_size_type), n, 0),
                    Zr = dplyr::if_else(grepl("correlation", effect_size_type), n, 0),
                 Other = dplyr::if_else(grepl("other|Q10", effect_size_type), n, 0),
                 Mean = dplyr::if_else(grepl("mean/proportion/raw", effect_size_type), n, 0),
                 lnCVR = dplyr::if_else(grepl("lnCVR", effect_size_type), n, 0),
                 Slope = dplyr::if_else(grepl("slope", effect_size_type), n, 0)) %>% 
  colSums(.) %>% data.frame(effect = names(.), n = .) %>% mutate(prop = n / sum(n)) %>% arrange(effect) %>% 
  mutate(effect = fct_relevel(effect, c("Other", "Zr", "lnRR", "SMD", "lnCVR", "Slope", "Mean"))) %>% 
  ggplot() + tm + 
  geom_col(aes(x = effect, y = prop), width = 0.5, fill = "black", colour = text.colour, size = 0.5) +
  scale_y_continuous(expand=c(0,0), limits = c(0,0.35)) +
  labs(x = "Effect Size",
       y = "Frequency") +
  coord_flip()
